---
title: "HW 1"
author: Your name here
output: html_document
---

```{r knitr_options , include=FALSE}
#Here's some preamble, which makes ensures your figures aren't too big
knitr::opts_chunk$set(fig.width=6, fig.height=3.6, warning=FALSE,
message=FALSE)
library(mosaic)
```

For this lab, we will be using the `mlb11` data set, as per Lab 1. Please submit a completed HTML file, due at the beginning of class. 

## Question 1

Write code to upload the `mlb11` data set into `R`, and print the first 6 rows of the data set.

```{r}
#Your code goes here

```





## Question 2

What element of the `mlb11` data set corresponds to the 10th row and 6th column? What team and variable is referred to in this cell?

## Question 3

Describe the distribution of at-bats (`at_bats`) by considering its center, shape, and spread. Back up your claims by identifying the mean, median, standard deviation and interquartile range of this variable.

## Question 4

A coach is curious if stealing more bases is linked with more home runs. Make and describe a scatter plot of team home runs versus stolen bases. Then, add a title to your plot.

## Question 5.

Create a new variable, `AboveAveWins`, representing whether or not the team won more than half of its games (there are 162 games in a season).  How many teams won 82 games or more?

## Question 6. 

You can create separate scatter plots within groups using the code below.

```{r}
xyplot(homeruns ~ stolen_bases | AboveAveWins, data = mlb11)
```

Does the association between home runs and stolen bases differ within the groups of `AboveAveWins`?

## Question 7. 

Make both a histogram and a boxplot of `hits`. What features are apparent in the histogram that aren't apparent in the boxplot? What features are apparent in the boxplot that aren't apparent in the histogram?


