---
title: "HW 3"
author: "Your name here"
output: pdf_document
---

```{r knitr_options , include=FALSE}
#Here's some preamble, which makes ensures your figures aren't too big
knitr::opts_chunk$set(fig.width=6, fig.height=3.6, warning=FALSE,
message=FALSE)
library(mosaic)
library(Lahman)
library(corrplot)
data(Teams)
data(Pitching)
data(Batting)
```

For this homework, we will be using data provided in the `Lahman` package. 

# Part 1


Our first goal is to use variables in the `Teams` data set to identify the model that best fits the number of runs scored for a team in a single season. We'll use all seasons since 1970. Several models are proposed. 

```{r}
Teams.1 <- filter(Teams, yearID >=1970)
Teams.1 <- mutate(Teams.1, X1B = H - X2B - X3B - HR)

fit.1 <- lm(R ~ X1B + X2B + X3B + HR, data = Teams.1)
fit.2 <- lm(R ~ X1B + X2B + X3B + HR + BB, data = Teams.1)
fit.3 <- lm(R ~ X1B + X2B + X3B + HR + BB + SO, data = Teams.1)
fit.4 <- lm(R ~ X1B + X2B + X3B + HR + BB + SO + CS, data = Teams.1)
fit.5 <- lm(R ~ X1B + X2B + X3B + HR + BB + SO + CS + lgID, data = Teams.1)
fit.6 <- lm(R ~ X1B + X2B + X3B + HR + BB + SO + CS + lgID + SB, data = Teams.1)

options(scipen=999)
```

Note: The `options(scipen = 999)` command disables R's scientific notation.

## Question 1

Using the AIC criteria, which of the six models would you recommend for measuring runs scored on a team-wide level?  From a baseball perspective, what does your choice suggest about certain measurements as far as their link to runs scored?


## Question 2

One of the coefficients in `fit.5` and `fit.6` is `lgID`. Generate a table of the `lgID` in your data set. What does this variable refer to?


## Question 3

Using the code below, the coefficient for `league = "NL"` is negative. Interpret this coefficient. What about baseball's rules make it important to consider which league each team played in? Note: you can google the differences between the American League and the National League to guide you.

```{r, eval = FALSE}
summary(fit.5)
```


## Question 4

Interpret the R-squared from `fit.5`, and produce model checks to determine if the assumptions for linear regression are appropriate. 


## Question 5

The first team in the data set is the Atlanta Braves, who scored 736 runs in 1971. Using `fit.5`, estimate how many runs your model expected the Braves to score. Did the Braves outperform expectations (score more runs) or underperform expectations (score fewer runs)? 


# Part II

In this part, we'll use the `Hitting` data set, and explore the properties of runs created. In this code, we look at three formulas for runs created: `RC1`, `RC2`, and `RC3`. 

```{r}
Batting.1 <- Batting %>% 
 filter(yearID >=1971, AB > 500) %>%
 mutate(X1B = H - X2B - X3B - HR, 
        TB = X1B + 2*X2B + 3*X3B + 4*HR, 
        RC1 = (H + BB)*TB/(AB + BB),
        RC2 = (H + BB - CS)*(TB + (0.55*SB))/(AB+BB),
        RC3 = ((H + BB - CS + HBP - GIDP)*(TB + (0.26*(BB - IBB + HBP))) + 
          (0.52*(SH+SF+SB)))/(AB+BB+HBP+SH+SF))

Batting.2 <- Batting.1 %>%
  arrange(playerID, yearID) %>%
  group_by(playerID) %>%
  mutate(f.RC1 = lead(RC1), f.RC2 = lead(RC2), f.RC3 = lead(RC3)) %>%
  na.omit()

cor.matrix <- cor(select(ungroup(Batting.2), 
    f.RC1, f.RC2, f.RC3, RC1, RC2, RC3), 
    use="pairwise.complete.obs")
corrplot(cor.matrix, method = "number")

```

## Question 6

Which of the three runs created metrics more strongly correlates with its own future performance in the following year?  Is that a good thing or a bad thing?


## Question 7

Which of the three runs created metrics more strongly correlates with `f.RC1`? What does this suggest?


## Question 8

Make a pair of scatter plots: `f.RC1` versus `RC1`, and `f.RC3` versus `RC3`. Is the difference in correlations between each pair of variables obvious in the scatter plots?

## Question 9

In place of correlation, a useful tool for measuring predictive accuracy is root mean squared error (MAE). In R, the MAE between two variables of the same length can be calculated as follows:

```{r}
x <- c(6, 10, 10)
y <- c(10, 10, 8)
MAE <- mean(abs(x-y))
MAE
```

Calculate the MAE between between each of the following three variable pairs: `RC1` and `f.RC1`, `RC2` and `f.RC2`, and `RC3` and `f.RC3`.

## Question 10

Interpret the MAE between `RC1` and `f.RC1`. Is there a noticeable difference between your MAE's found in question 9?  What does this suggest about the more complicated `RC3` formula?


