---
title: "Homework 7"
output:
  pdf_document: default
  html_document:
    css: ../lab.css
    highlight: pygments
    theme: cerulean
author: MA 276, Skidmore College
---

## Overview

In this lab, we'll practice implementing logistic regression to estimate the probability of successful NBA shots. We'll also link to shot-level probabilities and expected points. Before we do anything, we have to load and clean the data, as in Lab 6. 

```{r, eval = FALSE}
library(RCurl)
library(mosaic)
url <- getURL("https://raw.githubusercontent.com/JunWorks/NBAstat/master/shot.csv")
nba.shot <- read.csv(text = url)
nba.shot <- na.omit(nba.shot)
nba.shot <- filter(nba.shot, PTS <4, SHOT_DIST>=22 |PTS_TYPE==2)
nrow(nba.shot)
```



\section{Expected points}

All else being equal, what's the most efficient shot in the NBA?

In our lab, we characterized by points type using the following code:

```{r, eval = FALSE}
tally(SHOT_RESULT ~ PTS_TYPE, data = nba.shot, format = "proportion")
```

Of course, all two-point shots are not created equal. Using the cut command, we split two-pointers by distance into different groups, labeled `D1` to `D8`, in order from shortest to longest and grouped by shot type (2 or 3 points). The two data sets, `nba.two` and `nba.three` contain the two and three-pointers, respectively.

```{r, eval  = FALSE}
nba.two <- nba.shot %>%
  filter(PTS_TYPE == 2) %>%
  mutate(dist.cat = cut(SHOT_DIST, breaks = c(-100, 3, 6, 12, 100), 
                        labels = c("D1", "D2", "D3", "D4")))
  
nba.three <- nba.shot %>%
  filter(PTS_TYPE == 3) %>%
  mutate(dist.cat = cut(SHOT_DIST, breaks = c(0, 23, 25, 100), 
                        labels = c("D6", "D7", "D8")))  


tally(SHOT_RESULT ~ dist.cat, data = nba.two, format = "proportion")
tally(SHOT_RESULT ~ dist.cat, data = nba.three, format = "proportion")
```

### Question 1

In order from best (highest expected points) to worst (lowest), order the categories D1 to D8.


### Question 2

Using code from our last lab, identify of expected points are higher on two or three point shots taken by Rajon Rondo. 


### Question 3

Here's are two models of shot success (note that we re-bind all of the shots together).

```{r, eval = FALSE}
nba.shot2 <- rbind(nba.two, nba.three)

fit.1 <- glm(SHOT_RESULT == "made" ~ SHOT_DIST + TOUCH_TIME + 
            DRIBBLES + SHOT_CLOCK + CLOSE_DEF_DIST, 
             data = nba.shot2, family = "binomial")

fit.2 <- glm(SHOT_RESULT == "made" ~ dist.cat + TOUCH_TIME + 
            DRIBBLES + SHOT_CLOCK + CLOSE_DEF_DIST, 
             data = nba.shot2, family = "binomial")

```

Using the AIC criteria, which is the preferred fit of shot success? Is it close? 


### Question 4

Using `fit.2`, estimate the increased odds of a made shot given a one-unit increase in closest defender distance. Then, estimate the increased odds of a made shot given a ten-unit increase in closest defender distance.


### Question 5

Add game location (`LOCATION`) to `fit.2`. Does this improve the fit? Is the coefficient for this term statistically and/or practically significant?  What does that suggest? 


### Question 6

Does it make sense to add if the shooter's team was victorious (variable `W`) or margin of victory (`FINAL_MARGIN`) to the model? Why or why not? You do not need to run any code to answer this. 


### Question 7

Using Seth's article [here](https://sports.vice.com/en_us/article/moreyball-goodharts-law-and-the-limits-of-analytics) and referencing the charts shown, explain Goodhart's law as it applies to statistics in the NBA. 
