---
title: "HW 8"
author: "Your name here"
output: pdf_document
---

```{r knitr_options , include=FALSE}
#Here's some preamble, which makes ensures your figures aren't too big
knitr::opts_chunk$set(fig.width=6, fig.height=3.6, warning=FALSE,
message=FALSE, eval = FALSE)
```

## Introduction

In this assignment, we'll return to Stein's Paradox, except this time through the lens of batting averages and on base percentages in baseball. Recall, batting average is a batters hits divided by at bats, and on base percentage is the frequency of times a batter reached base divided by at bats. 

The `Batting` data set contains player-season information for each hitter since 1971. For obtaining a James-Stein estimator, it doesn't make sense to compare players across several different era's, so we'll focus more on recent players. We also create new variables for batting average (`BA`) and on base percentage (`OBP`).

```{r}
library(mosaic)
library(Lahman)
library(corrplot)
data(Batting)
Batting.1 <- Batting %>%
  filter(yearID >= 2000) %>%
  mutate(BA = H/AB, OBP = (H+BB+HBP)/AB) %>%
  na.omit()
head(Batting.1)
```

Let's go back to the year 2007 and identify players that had roughly 450 at bats. 

```{r}
first.players <- filter(Batting.1, yearID == 2007, AB >=445, AB <=455)
head(first.players)
```

There are 9 players in this data set.

1. For batting average (`BA`), calculate the James-Stein estimate of the nine players career batting percentages using the `Batting.02` data as a base. A few hints can be provided below. 

**Hint 1**: The estimate of the variance term comes from the binomial distribution. As in Lab 9, this can be estimated using the following code

```{r}
sigma.sq <- p.bar*(1-p.bar)/450 ##Rough approximation
sigma.sq
```

**Hint 2**: Here's a data set that extracts players career batting averages using all years since 2002. 

```{r}
all.players <- Batting.1 %>%
  group_by(playerID) %>%
  filter(playerID %in% first.players$playerID, yearID >= 2002) %>%
  summarise(BA.career = sum(H)/sum(AB), OBP.career = sum(H+AB+HBP)/sum(AB))
```

You can link the `all.players` and `first.players` data sets by using `inner_join()`: See Lab 9. 


2. What is the relative amount of shrinkage towards the overall league average that we can expect for players' batting average? 


3. Refer to the original Efron and Morris paper. Does this amount of shrinkage surprise you given their findings and their sample size?


4. Compare the root mean squared error of the J-S estimate and the MLE estimate (recall, the MLE is the players original batting average). Which is more accurate?


5. One of the players is Miguel Olivo (`playerID` = `olivomi01`), a catcher who batted 0.240 for his career. Catchers generally tend to hit worse than other position players. Explain how this may be impacting the accuracy of the J-S estimate.


6. Let's fix a players number of at bats. Given what we know about on base percentage and batting average, for which variable do you expect a larger reversion towards the overall league-wide average? 


7. *Extra credit*: Repeat the same analysis, except for OBP. Do your findings match your expectations from Question 6?  For OBP, is the J-S estimate more accurate than the MLE estimate?


