---
title: "HW 7: NHL stats"
author: Stats and sports class
date: Fall 2019
output: 
  pdf_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, eval = FALSE, digits = 3, warning = FALSE)
```

# Preliminary notes for doing HW

1. All files should be knit and compiled using R Markdown. Knit early and often! I do not recommend waiting until the end of the HW to knit.

2. All questions should be answered completely, and, wherever applicable, code should be included. 

3. If you work with a partner or group, please write the names of your teammates. 
4. Copying and pasting of code is a violation of the Skidmore honor code

# Homework questions

## Part I: Readings

1. Read the summary model by the Evolving Wild twins: 

[https://rpubs.com/evolvingwild/395136/](https://rpubs.com/evolvingwild/395136/)

Describe five unique hockey features that were implemented in their model. That is, look through their code, and highlight various ways that hockey-specific knowledge changed how they approached the problem.

2. Compare the three variable importance plots. Which variables were more important during even-strength play? Which were more important (relatively speaking) when a team was shorthanded or at uneven strength?


## Part II: Implementation

We can access recent shot data here:

```{r}
library(RCurl); library(tidyverse)
gitURL<- "https://raw.githubusercontent.com/statsbylopez/StatsSports/master/Data/pbp_data_hockey.rds"
pbp_data <- readRDS(gzcon(url(gitURL)))
names(pbp_data)
dim(pbp_data)
```


## Question 1

Create a new variable for whether or not the shot occured during 5 v 5 play (that is, `home_skaters==5` and `away_skaters` == 5). Call this variable `is_5v5`. 

Next, identify the goal rate (e.g, how often each shot was turned into a goal) within each cohort of `is_5v5`. That is, were shots more or less likely to go in during 5v5 play?

## Question 2

Run the model below

```{r}
library(broom)
fit_1 <- glm(event_type == "GOAL" ~ event_distance + 
               event_angle + event_detail ,
             family = "binomial", data = pbp_data)
tidy(fit_1)
```

Interpret the coefficient on `event_detailWrist`

## Question 3

Add `is_5v5` to your model in Question 2. Using AIC criterion, identify if this creates a preferable model.

## Question 4

For `game_id == 2017020324`, identify each participating team's goals and expected goals. Did the outcome of this game match the relative shot inputs?

## Bonus

Find the one game across the last two seasons where the different between the observed goal differential was as different from the expected goal differential
