---
title: "HW 9: Soccer stats"
author: Stats and sports class
date: Fall 2019
output: 
  pdf_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, eval = TRUE, digits = 3, warning = FALSE)
```


## Overview

In this HW, we'll look at women's world cup data (note -- several questions are similar to the most recent lab).

```{r, eval = TRUE}
library(RCurl)
library(tidyverse)
url <- getURL("https://raw.githubusercontent.com/statsbylopez/StatsSports/master/Data/sb_shot_data.csv")
wwc_shot <- read.csv(text = url)
```


## Better shot maps

`ggplot()` has ample ways to enhance shot maps. Consider the following maps

```{r, eval = TRUE}
wwc_shot <- wwc_shot %>% 
  mutate(is_goal = shot.outcome.name == "Goal")

usa_shot <- wwc_shot %>% 
  filter(possession_team.name == "United States Women's")


p1 <- ggplot(usa_shot, aes(location.x, location.y)) + 
  geom_point() 

p2 <- ggplot(usa_shot, aes(location.x, location.y, colour = is_goal)) + 
  geom_point() 

p3 <- ggplot(usa_shot, aes(location.x, location.y, 
      colour = is_goal, size = shot.statsbomb_xg)) + 
  geom_point() 

p1
p2
p3
  
```

1. What features are apparent in `p2` that aren't apparent in `p1`?  What features are apparent in `p3` that aren't apparent in `p2`.

**`p2` highlights where the actual goals were scored (generally closer to the net) that `p1` does not. `p3` gets at the size of the expected goal (how likely each show was of scoring) and features both distance and other shot level information**




## Practice with dplyr

8. For each USA shooter, average the `TimeInPoss` and `DefendersBehindBall` when they took their shot. Filter to make sure you are only looking at players with at least 10 shots. What does this say about how players took shots?

```{r, eval = TRUE}
usa_shot %>% 
  group_by(player.name) %>% 
  summarise(mean_time = mean(TimeInPoss), 
            mean_def = mean(DefendersBehindBall), 
            n_shots = n()) %>% 
  filter(n_shots >= 10) 
```


**Julie Ertz tended to take shots quicker, while Alex Morgan and Roze Lavelle's took longer to build up. Ertz has the *most* defenders in front of her, while Carli Lloyd and Megan Rapinoe had the fewest**



## Exploration

A soccer coach wants to know the best places to shoot from. What would you tell the coach?  Create a grid across the field using the `cut()` command (for both x and y), and then, within each location, estimate the goal rate. Next, use `geom_tile()` to make a map of goal rates within each cell of the grid you created. For a reminder on `cut()`, see our notes on Hosmer-Lemeshow, or `?cut()`.

```{r}
wwc_shot %>% 
  mutate(cut_x = cut(location.x, 10), 
         cut_y = cut(location.y, 10)) %>% 
  group_by(cut_x, cut_y) %>% 
  summarise(goal_rate = mean(is_goal)) %>% 
  ggplot(aes(cut_x, cut_y, fill = goal_rate)) + 
  geom_tile()
```

**Solutions: The above map creates a grid and plot that is useful for assesing goal likelihood -- students that are closest (in idea, not neccessarily in the exact visual) will get full credit**


