---
title: "HW 9: Soccer stats"
author: Stats and sports class
date: Fall 2019
output: 
  pdf_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, eval = FALSE, digits = 3, warning = FALSE)
```


## Overview

In this HW, we'll look at women's world cup data (note -- several questions are similar to the most recent lab).

```{r, eval = FALSE}
library(RCurl)
library(tidyverse)
url <- getURL("https://raw.githubusercontent.com/statsbylopez/StatsSports/master/Data/sb_shot_data.csv")
wwc_shot <- read.csv(text = url)
names(wwc_shot)
```

## Readings

Read xCommentary at [https://statsbomb.com/2016/10/xcommentary/](https://statsbomb.com/2016/10/xcommentary/). 

1. "We have commentary that doesn't understand how the game actually works." What specific example in soccer is he referring to?

2. "Instead we get "distance run" stats, which to my knowledge have never been proven as relevant to anything in football" -- Provide an example of another sport where new or different statistics has been provided, despite there being no obvious relevance to winning.

## Better shot maps

`ggplot()` has ample ways to enhance shot maps. Consider the following maps

```{r, eval = FALSE}
wwc_shot <- wwc_shot %>% 
  mutate(is_goal = shot.outcome.name == "Goal")

usa_shot <- wwc_shot %>% 
  filter(possession_team.name == "United States Women's")


p1 <- ggplot(usa_shot, aes(location.x, location.y)) + 
  geom_point() 

p2 <- ggplot(usa_shot, aes(location.x, location.y, colour = is_goal)) + 
  geom_point() 

p3 <- ggplot(usa_shot, aes(location.x, location.y, 
      colour = is_goal, size = shot.statsbomb_xg)) + 
  geom_point() 
  
```

1. What features are apparent in `p2` that aren't apparent in `p1`?  What features are apparent in `p3` that aren't apparent in `p2`.

2. The following contour plot creates lines where the team has shot in highest densities. The inside line is most `dense`, corresponding to the center of where the team took shots. What features are apparent in `p4` that aren't apparent in `p3`? What is apparent in `p3` that isn't in `p4`?

```{r, eval = FALSE}
p4 <- usa_shot %>% 
  ggplot(aes(location.x, location.y)) + 
  stat_density_2d()
p4
```


3. Find another team `wwc_shot %>% count(possession_team.name)` and plot their shots. How do they compare to the USA Women's team?

## Goals versus expectation

4. Let's investigate finishing ability on the USA team. Calculate the total number of goals scored by each player. Who actually scored the most goals?

5. Calculate the number of expected goals scored by each player. Who was expected to score the most goals?

6. The code below (using `group_by(), summarise(), mutate()`), calculate the performance above/below expectation for each member of the USA team who took a shot. Who performed better than expectation? Below?  What does the overall distribution say about the USA team?

```{r, eval = FALSE}
usa_shot %>% 
  group_by(player.name) %>% 
  summarise(xg = sum(shot.statsbomb_xg), 
            g = sum(is_goal), 
            n_shots = n()) %>% 
  mutate(ou = g-xg) %>% 
  arrange(ou)
```

7. Annotate each line of code above to identify what it's doing. 

## Practice with dplyr

8. For each USA shooter, average the `TimeInPoss` and `DefendersBehindBall` when they took their shot. Filter to make sure you are only looking at players with at least 10 shots. What does this say about how players took shots?

9. Among all players `wwc_shot`, identify the player who finished with the *most* and *least* goals above expectation.

10. Among all goalies (`player.name.GK`), identify the goalie who finished with the *most* and *least* goals allowed above expectation. 

11. Among all players, identify the player who took the most headers (`shot.body_part.name == "Head"`).

12. Among all players, estimate the goal rate given different `shot.technique.name`. Which of these tends to lead to have the highest chance of success?

13. Compare the average shot distance by different play patterns (`play_pattern.name`) -- which plays tend to lead to shots from longer distances? Shorter distances?

14. Make a goalie map -- that is, find a goalie, and make a shot map (most similar to `p3` above) that shows how the goalie fared in this tournament. Pick any goalie you want!

## Exploration

A soccer coach wants to know the best places to shoot from. What would you tell the coach?  Create a grid across the field using the `cut()` command (for both x and y), and then, within each location, estimate the goal rate. Next, use `geom_tile()` to make a map of goal rates within each cell of the grid you created. For a reminder on `cut()`, see our notes on Hosmer-Lemeshow, or `?cut()`.




